Decoding UK Hospital Wait Times: Systemic Challenges Beyond Resources

### 2024-11-28 paper/paper.qmd
Added literature, final references, and completed model.

---
title: "Decoding UK Hospital Wait Times: Infrastructure, Workforce, and Demand"
subtitle: "How Systemic Pressures Shape Delays in Medical Care"
author: 
  - Chenika Bukes
thanks: "Code and data are available at: [https://github.com/chenikabukes/UnitedKingdomHealthcare](https://github.com/chenikabukes/UnitedKingdomHealthcare)."
date: today
date-format: long
abstract: "Hospital wait times in the United Kingdom are a critical indicator of healthcare efficiency, reflecting the interplay between resources and patient demand. This study investigates how hospital infrastructure, workforce availability, and patient attendance rates influence wait times for fifteen key medical procedures between 2015 and 2022. Regression modeling reveals no statistically significant predictors, with hospital bed availability, physician density, and attendance rates showing weak associations with wait times. These findings suggest that systemic inefficiencies and external disruptions, such as the COVID-19 pandemic, play a more significant role than individual resource metrics. This study underscores the need for a deeper understanding of healthcare delays and highlights the importance of data-driven strategies to improve resource allocation and patient outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Timely access to healthcare is fundamental to achieving optimal health outcomes. In the United Kingdom, the National Health Service (NHS) faces mounting pressure to manage increasing patient demands amidst resource constraints. Hospital wait times, a key performance indicator, have become a focal point for policymakers and stakeholders alike. Extended wait times not only compromise patient care but also signal underlying inefficiencies in healthcare delivery systems.

This paper investigates how hospital infrastructure, workforce availability, and patient demand influence wait times for medical procedures across the UK. Specifically, it examines the roles of hospital beds per capita, physicians per 1,000 inhabitants, and Type 1 attendance rates in shaping delays for procedures like knee replacements, coronary artery bypass grafts, and cataract surgeries. While previous studies have explored individual contributors to wait times, this analysis integrates multiple predictors to provide a holistic view of systemic pressures.

The estimand in this study is the average percentage change in wait times for seven key medical procedures: knee replacements, hip replacements, prostatectomy, hysterectomy, coronary artery bypass grafts, transluminal coronary angioplasty, and cataract surgeries in the UK. The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. This measure quantifies how changes in healthcare infrastructure and demand impact system-wide delays, offering insights into the relative contribution of each predictor to observed wait times.

@@

In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR], using the haven and `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al. 2021). All figures in the report are generated using `ggplot2` [@citeGG] and `patchwork` [@patchwork]. We run the model in R [@citeR] using the `modelsummary` package of [@modelsummary].

This report integrates data from two key sources: the NHS and the OECD. These datasets provide critical information to analyze trends in healthcare demand and resources in the United Kingdom.

The NHS Referral to Treatment (RTT) waiting times dataset [@wait] is the primary source for waiting time statistics. These data measure the median time (in weeks) between referral by a general practitioner (GP) and the start of treatment for elective care, capturing delays across 15 key medical procedures such as general surgery, urology, and trauma & orthopedics. This dataset offers procedure-specific insights that allow for cross-procedure and temporal comparisons.

The NHS A&E Attendances and Emergency Admissions dataset [@attendance] captures the monthly demand for emergency services. It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission. These statistics are collected at the provider organization level from NHS Trusts, Foundation Trusts, and independent sector organizations. The data are aggregated from monthly submissions, a system adopted following Sir Bruce Keogh’s review of waiting time standards, and include attendance and admission figures.

Additional datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data on healthcare systems. For the United Kingdom, the OECD compiles data on hospital resources hospital beds and physician counts from NHS Digital and Public Health Scotland. The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments, making it a reliable complement to NHS data for the analysis.

## Measurement
**Physicians per 1,000** [@physicians]. Physician counts are sourced from the OECD, relying on NHS Digital, Public Health Scotland, and the General Medical Council (GMC). The data reflect licensed physicians, encompassing both general practitioners and specialists. The metric is calculated as the total number of physicians per 1,000 population. The dataset includes both headcount and rolecount metrics, with post-2009 data transitioning to headcount for greater accuracy. Physician counts include GP retainers and full-time equivalents, providing an accurate picture of the available workforce. Historical data adjustments account for changes in collection methodologies.  The data are collected and reported monthly through NHS administrative systems, with automated validation checks ensuring accuracy and consistency. Similar datasets include WHO and Eurostat health statistics offer broader regional comparisons but lack the granularity needed for a UK-focused study. [@https://turasdata.nes.nhs.scot/about-our-data-and-reports/methods/nhsscotland-workforce-data-warehouse/]

**Beds per 1,000 Population** [@beds]: This metric, provided by the OECD, uses data from NHS Digital, Public Health Scotland, and national agencies across the UK. It tracks the availability of inpatient beds. Annual averages of beds available overnight in public hospitals. Includes acute care and psychiatric beds but excludes private sector facilities for consistency. Data are for financial years and represent publicly funded healthcare infrastructure. No significant methodological breaks are noted for this variable.

@@

**Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors. While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2019. Similar datasets from the World Health Organization (WHO) could offer supplementary insights but were not utilized due to a lack of UK-specific detail.

## Methodology {#sec-methodology}
The data cleaning method is described in @sec-cleaning. All data cleaning was done using `arrow` [@arrow], `tidyverse` [@tidyverse], `readxl` [@readxl], and `stringr` [@stringr]. Below is a snapshot of the first five rows of the key data columns which will be used for this paper. @tbl-data-appendix provides the entire cleaned dataset in table format.

@@

# Display the first part of the table
kable(head(part1), 
      caption = "Part 1: Hospital statistics")
```

@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of 10 variables and 8 observations spanning the years 2015 to 2022. The variables include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. Additional variables capture wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery. In @tbl-data we see the average percentage change in wait times from 2015 levels in the `Total` column. All metrics are standardized based on population or healthcare utilization data for each year.

@@

@fig-data depicts the percentage increase in wait times for various medical procedures compared to 2015. By 2019, most procedures show a notable increase in wait times, with angioplasty and prostatectomy exhibiting the largest rises. While some procedures, such as knee and hip surgeries, show consistent increases, others, like prostatectomy, initially decline before rising sharply. The steady increase across most procedures highlights a systemic trend of growing delays in accessing medical care, potentially signaling resource limitations or increased demand in healthcare services. Overall, wait times have risen by over 20% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.

@@

@fig-total shows the percentage increase in total wait times for key medical procedures in the UK relative to 2015, revealing a steady rise through 2019, followed by fluctuations during the COVID-19 pandemic. The moderate growth between 2015 and 2019 suggests manageable strain on healthcare resources, but the dip in 2020 reflects reduced elective procedures as hospitals prioritized emergency care. This was followed by a sharp increase in 2021 and a steep surge in 2022, likely due to the backlog created during the pandemic, ongoing demand pressures, and resource constraints. These trends underscore the urgent need for targeted healthcare interventions, including expanded capacity, streamlined workflows, and policies to address the post-pandemic backlog and increasing demand.
## Predictor variables

@@

@@

\begin{align} 
Y_{ij} & = \text{The average percentage change in wait time for the 15 specific medical procedures in year i}. \\
X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \
\beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
\epsilon_{i} & = \text{The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.}
\end{align}

@@

Each predictor variable is expected to contribute uniquely:

- Beds per 1,000 inhabitants (X_1): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
- Physicians per 1,000 inhabitants (X_2): Greater physician availability is hypothesized to decrease wait times.
- Attendance Rates (X_3): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.

We run the model in R [@citeR] using the `modelsummary` package of `rsaratnam` [@rsaratnam].

@@

@@

```
The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (Pct_Wait_Avg) compared to three key predictors: attendance, physicians_per_1000, and beds_per_1000. The red lines depict the predicted relationships, while the blue points represent the observed data values. The attendance predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend. Physicians_per_1000 also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance. In contrast, beds_per_1000 shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend. These plots suggest that attendance is the most influential predictor of wait times among the three variables, while beds_per_1000 has the least impact.

@fig-modeldetails presents the residual plots. The residual plots indicate that the linear model generally fits the data well, with residuals scattered randomly around zero for all predictors, suggesting no major violations of linearity or independence assumptions. However, in each residual plot there is a violation of constant variance where there are noticeable outliers and potential influential points, particularly in the plots for attendance and physicians per 1,000, which may affect the model's robustness. The fitted vs. actual values plot shows a good overall alignment with most points near the predicted line, but two significant deviations highlight discrepancies that could stem from these outliers. Given the small sample size, these findings should be interpreted with caution.

@@

## Overview
This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for key medical procedures in the United Kingdom from 2015 to 2022. Our analysis employs a linear regression model but finds no statistically significant relationships between the predictors and wait times. While hospital beds per 1,000 inhabitants showed a weak positive association with wait times, the relationship was not significant, suggesting that bed capacity alone may not directly mitigate delays. Similarly, physician density and attendance rates demonstrated positive but insignificant associations with wait times, reflecting the complex interplay of factors influencing healthcare delays. These findings indicate that systemic inefficiencies, resource allocation challenges, and unmeasured variables may play a more significant role than initially hypothesized. The limited explanatory power of the model and the presence of anomalies in the data post-2019 highlight the need for more detailed data and refined analyses to uncover the true drivers of healthcare delays.

@@

```{r}
#| echo: false
#| warning: false
#| error: false
#| label: tbl-data-appendix
#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
# Import the cleaned healthcare dataset
cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

@@

# Display the first part of the table
kable(head(part1), 
      caption = "Part 1: Hospital statistics")
kable(head(part2), 
      caption = "Part 2: Hospital statistics")
kable(head(part3), 
      caption = "Part 3: Hospital statistics")
kable(head(part4), 
      caption = "Part 4: Hospital statistics")
```

@@

print(plot3)
```
## Model Details {#sec-model-details}
### Original Model Residuals
```{r}
#| echo: false
#| output: asis
#| eval: true
#| label: fig-modeldetails
#| tbl-cap: "Residual plots of each predictor."
#| # Load necessary libraries
library(ggplot2)

@@

# Print the combined plot
print(combined_plot)
```
### Adapted Model {#sec-adapted}
```{r}
#| echo: false
#| output: asis
#| eval: true
#| label: tbl-modeladj
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors"

#### Workspace setup ####
library(tidyverse)

#### Read data ####
# Load the cleaned data
analysis_data <- read_csv("../data/02-analysis_data/final_healthcare_data.csv")

#### Filter data ####
# Exclude data from 2020 to 2022
filtered_data <- analysis_data %>% 
  filter(Year < 2020)

@@

# Display the model summary in a table
modelsummary::modelsummary(
  list("Frequentist Linear Model" = modela),
  statistic = c("p.value")
)

>>>>

---
title: "Decoding UK Hospital Wait Times: Systemic Challenges Beyond Resources"
subtitle: "How Infrastructure, Workforce, and Demand do not Explain Delays in Medical Care"
author: 
  - Chenika Bukes
thanks: "Code and data are available at: [https://github.com/chenikabukes/UnitedKingdomHealthcare](https://github.com/chenikabukes/UnitedKingdomHealthcare)."
date: today
date-format: long
abstract: "Hospital wait times in the United Kingdom are a critical indicator of healthcare efficiency, reflecting the interplay between resources and patient demand. This study investigates how hospital infrastructure, workforce availability, and patient attendance rates influence wait times for fifteen key medical procedures between 2015 and 2022. Regression modeling reveals no statistically significant predictors, with hospital bed availability, physician density, and attendance rates showing weak associations with wait times. These findings suggest that systemic inefficiencies and external disruptions, such as the COVID-19 pandemic, play a more significant role than individual resource metrics. This study underscores the need for a deeper understanding of healthcare delays and highlights the importance of data-driven strategies to improve resource allocation and patient outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Timely access to healthcare is fundamental to achieving optimal health outcomes. In the United Kingdom, the National Health Service (NHS) faces mounting pressure to manage increasing patient demands amidst resource constraints. Hospital wait times, a key performance indicator, have become a focal point for policymakers and stakeholders alike [@health]. Extended wait times not only compromise patient care but also signal underlying inefficiencies in healthcare delivery systems.

This paper investigates how hospital infrastructure, workforce availability, and patient demand influence wait times for medical procedures across the UK. Specifically, it examines the roles of hospital beds per capita, physicians per 1,000 inhabitants, and Type 1 major care attendance rates in shaping delays from 2015 to 2022 for fifteen major fields of medicine presented in @tbl-treatments. The NHS consistently met the 18-week standard from 2008 to 2015, but has not since, thus the years 2015 on wards were selected for investigation. NHS waiting times continue to sharply increase making this a national healthcare crisis in the UK [@lancet].

The estimand in this study is the average percentage change from 2015 in wait times for fifteen key fields of medicine @tbl-treatments. The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. This measure quantifies how changes in healthcare infrastructure and demand impact system-wide delays, offering insights into the relative contribution of each predictor to observed wait times.

@@

In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR] and use the `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al. 2021) packages. All figures in the report are generated using `ggplot2` [@citeGG] with `patchwork` [@patchwork], and `gridExtra` [@gridExtra].

This report integrates data from two key sources: the NHS and the Organisation for Economic Co-operation and Development (OECD). These datasets provide critical information to analyze trends in healthcare demand and resources in the United Kingdom.

The NHS Referral to Treatment (RTT) waiting times dataset [@wait] is the primary source for waiting time statistics. These data measure the median time (in weeks) between referral by a general practitioner (GP) and the start of treatment for elective care, capturing delays across 15 key medical procedures @tbl-treatments. This dataset offers procedure-specific insights that allow for cross-procedure and temporal comparisons. The data are aggregated and published monthly.

The NHS A&E Attendances and Emergency Admissions dataset [@attendance] captures the monthly demand for emergency services. It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission. These statistics are collected at the provider organization level from NHS Trusts, Foundation Trusts, and independent sector organizations. The data are aggregated from monthly submissions.

Additional datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data on healthcare systems. For the United Kingdom, the OECD compiles data on hospital resources hospital beds and physician counts from NHS Digital and Public Health Scotland [@scotland]. The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments.

## Measurement
**Physicians per 1,000** [@physicians]. Physician counts are sourced from the OECD, relying on NHS Digital, Public Health Scotland, and the General Medical Council (GMC). The data reflect licensed physicians, encompassing both general practitioners and specialists. The metric is calculated as the total number of physicians per 1,000 population. The dataset includes both headcount and rolecount metrics, with post-2009 data transitioning to headcount for greater accuracy. Physician counts include GP retainers and full-time equivalents, providing an accurate picture of the available workforce. Historical data adjustments account for changes in collection methodologies. The data are collected and reported monthly through NHS administrative systems, with automated validation checks ensuring accuracy and consistency. The population data is taken from population estimates as those registered with the NHS.

**Beds per 1,000 Population** [@beds]: This metric, provided by the OECD, uses data from NHS Digital, Public Health Scotland, and national agencies across the UK. It tracks the availability of inpatient beds. It provides annual averages of beds available overnight in public hospitals. Includes acute care and psychiatric beds but excludes private sector facilities for consistency. Data are for financial years and represent publicly funded healthcare infrastructure. The number of beds data is gathered through automated administrative streams from NHS hospitals and care facilities. The population data is taken from population estimates as those registered with the NHS.

@@

**Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors. While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2022. Similar datasets from the World Health Organization (WHO) and Eurostat health statistics could offer supplementary insights but were not utilized due to a lack of UK-specific detail.

## Methodology {#sec-methodology}
The data cleaning method involved many steps and is described in @sec-cleaning. All data cleaning was done using `arrow` [@arrow], `tidyverse` [@tidyverse], `readxl` [@readxl], and `stringr` [@stringr]. Below is a snapshot of the first five rows of the key data columns which will be used for this paper. @tbl-data-appendix provides the entire cleaned dataset in table format.

@@

# Display the first part of the table
kable(part1, 
      caption = "NHS Hospital Statistics for 2015-2022")
```

@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of the 3 predictor variables and estimand and 8 observations spanning the years 2015 to 2022. The predictors include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. The outcome variable `Total` captures wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery. All metrics are standardized based on population or healthcare utilization data for each year. The full cleaned dataset is provided in @ tbl-data-appendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.

@@

@@

@fig-total shows the percentage increase in total wait times for key medical procedures in the UK relative to 2015, revealing a steady rise through 2019, followed by fluctuations during the COVID-19 pandemic. The moderate growth between 2015 and 2019 suggests manageable strain on healthcare resources, but the dip in 2020 reflects reduced elective procedures as hospitals prioritized emergency care. This was followed by a sharp increase in 2021 and a steep surge in 2022, likely due to the backlog created during the pandemic, ongoing demand pressures, and resource constraints. These trends underscore the urgent need for targeted healthcare interventions, including expanded capacity, streamlined workflows, and policies to address the post-pandemic backlog and increasing demand. Overall, wait times have risen close to 30% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.

## Predictor variables

@@

The model was run in `R` [@citeR] using the `modelsummary` package [@modelsummary].

@@

\begin{align} 
Y_{ij} & = \text{The average percentage change in wait time for the 15 specific medical procedures in year i}. \\
X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \\
\beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
\epsilon_{i} & = \text{The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.}
\end{align}

@@

### Model Assumptions {#sec-model-assumptions}
The linear regression model assumes:

* Linearity: The relationship between the predictors and response variable is linear.
* Independence: Observations are independent of each other.
* Homoscedasticity: The variance of the error term (\epsilon) is constant across all levels of the predictors.
* Normality: The residuals of the model are approximately normally distributed.

@@

### Hypotheses
Each predictor variable is expected to contribute uniquely:

* Beds per 1,000 inhabitants (X_1): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
* Physicians per 1,000 inhabitants (X_2): Greater physician availability is hypothesized to decrease wait times.
* Attendance Rates (X_3): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.

### Evaluation and Validation
Evaluation and Validation
To ensure the robustness of our findings, the model was validated through the following steps:

1. Residual Diagnostics: We assessed residual plots for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
2. Significance Testing: We conducted t-tests to assess whether each coefficient (\beta_0, \beta_1, \beta_2, \beta_3) is statistically significant at the 0.05 level @sec-results.
3. Goodness of Fit: The R-squared statistic was calculated to measure how well the model explains the variability in wait times @sec-results.
4. Alternative Models: a variant containing data only from 2015 to 2019 was tested to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit @sec-adapted. In both models, the p-values of the coefficients were insignificant so the model containing the larger data set (original model from 2015-2022) was used for final analysis.

@@

```
The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (`Total`) compared to the three predictors: attendance, physicians_per_1000, and beds_per_1000. The red lines depict the predicted relationships, while the blue points represent the observed data values. The physicians_per_1000 predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend. Attendance also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance. In contrast, beds_per_1000 shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend. For each plot of the predictor vs outcome variable there seems to be at least two data points which deviate significantly from the remaining points. These were hypothesized to be due to the outliers from trend witnessed in @sec-data stemming from the Covid-19 pandemic years. @sec-adapted developed and analysed the significance of an adapted model for years 2015-2019 before the pandemic to avoid such outlier effects in capturing the trend. The overall model significance was much greater and linear assumption violations in the residual plots were non-existent, however individual coefficients were still non-significant and the dataset on which this model was constructed was even smaller with 5 observations.

@@

## Overview
This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for 15 medical sectors in the United Kingdom from 2015 to 2022. Our analysis employs a linear regression model but finds no statistically significant relationships between the predictors and wait times. While hospital beds per 1,000 inhabitants showed a weak positive association with wait times, the relationship was not significant, suggesting that bed capacity alone may not directly mitigate delays. Similarly, physician density and attendance rates demonstrated positive but insignificant associations with wait times, reflecting the complex interplay of factors influencing healthcare delays. These findings indicate that systemic inefficiencies, resource allocation challenges, and unmeasured variables may play a more significant role than initially hypothesized. The limited explanatory power of the model and the presence of anomalies in the data post-2019 highlight the need for more detailed data and refined analyses to uncover the true drivers of healthcare delays.

@@

### Treatments {#sec-treatments}

```{r}
#| echo: false
#| warning: false
#| error: false
#| label: tbl-data-appendix
#| tbl-cap: "A summary table of the treatments"

# Import the cleaned healthcare dataset
cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

# Extract column names starting from the 5th column to the second last column
treatment_names <- colnames(cleaned_data)[5:(ncol(cleaned_data) - 1)]

# Display the treatment names in a table
library(knitr)
kable(
  data.frame("Treatments" = treatment_names),
  caption = "List of Medical Treatment Areas"
)
```

### Full Cleaned Dataset
```{r}
#| echo: false
#| warning: false
#| error: false
#| label: tbl-data-appendix
#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"

@@

# Display the first part of the table
kable(part1, 
      caption = "Part 1: Hospital statistics")
kable(part2, 
      caption = "Part 2: Hospital statistics")
kable(part3, 
      caption = "Part 3: Hospital statistics")
kable(part4, 
      caption = "Part 4: Hospital statistics")

@@

print(plot3)
```
## Model Details {#sec-model-details}
### Original Model Residuals {#sec-residuals}
```{r}
#| echo: false
#| output: false
#| eval: true
#| label: fig-modeldetails
#| tbl-cap: "Residual plots of each predictor."

@@

# Print the combined plot
print(combined_plot)
```
@fig-modeldetails presents the residual plots which indicate that the linear model generally fits the data well, with residuals scattered randomly around zero for all predictors, suggesting no major violations of linearity or independence assumptions. However, in each residual plot there is a violation of constant variance where there are noticeable outliers and potential influential points, particularly in the plots for attendance and physicians per 1,000, which may affect the model's robustness. The fitted vs. actual values plot shows a good overall alignment with most points near the predicted line, but two significant deviations highlight discrepancies that could stem from these outliers. Given the small sample size, these findings should be interpreted with caution. 
### Adapted Model {#sec-adapted}
```{r}
#| echo: false
#| output: asis
#| eval: true
#| label: tbl-adj
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"

#### Read data ####
# Load the cleaned data
analysis_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

#### Filter data ####
# Include only data from 2015 to 2019
filtered_data <- analysis_data %>% 
  filter(Year >= 2015 & Year <= 2019)

@@

# Display the model summary in a table
modelsummary::modelsummary(
  list("Frequentist Linear Model" = modela),
  statistic = c("p.value")
)
```
The linear regression model, summarized in @tbl-adj, examines the impact of healthcare system factors on wait times from 2015–2019. Hospital beds per 1,000 inhabitants (\beta_1 =−176.99) and physicians per 1,000 (\beta_2 =−78.19) both show negative associations with wait times, suggesting increased resources reduce delays, though neither effect is statistically significant (p >0.05). Attendance rates (\beta_3=−0.66) exhibit a minimal effect. The model explains 97.8% of the variability in wait times ($R^2$=0.978), though the adjusted $R^2$ of 0.911 reflects possible overfitting due to the small sample size (n = 5). While the model fits well, the lack of significant predictors highlights the need for further data to better capture drivers of wait times. The small RMSE of 0.58 indicates the linear trend is never far off from the actual wait times.

```{r}
#| echo: false
#| output: false
#| eval: true
#| label: fig-residadj
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"
# Extract residuals and fitted values from the model
filtered_data <- filtered_data %>%
  mutate(
    residuals = residuals(modela),
    fitted = fitted.values(modela)
  )

#### Diagnostic Plots ####

# Plot b: Residuals vs Predictor 1 (attendance)
plot_b <- ggplot(filtered_data, aes(x = attendance, y = residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
  theme_classic() +
  labs(
    title = "Residuals vs Attendance",
    y = "Residuals",
    x = "Attendance"
  )

# Plot c: Residuals vs Predictor 2 (physicians_per_1000)
plot_c <- ggplot(filtered_data, aes(x = physicians_per_1000, y = residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
  theme_classic() +
  labs(
    title = "Residuals vs Physicians per 1,000",
    y = "Residuals",
    x = "Physicians per 1,000"
  )

# Plot d: Fitted values vs Actual values (Total)
plot_e <- ggplot(filtered_data, aes(x = Total, y = fitted)) +
  geom_point(color = "blue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  theme_classic() +
  labs(
    title = "Fitted vs Actual Values",
    y = "Fitted Values",
    x = "Actual Values"
  )

# Plot e: Residuals vs Predictor 3 (beds_per_1000)
plot_d <- ggplot(filtered_data, aes(x = beds_per_1000, y = residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
  theme_classic() +
  labs(
    title = "Residuals vs Beds per 1,000",
    y = "Residuals",
    x = "Beds per 1,000"
  )

# Combine all plots into a grid
combined_plot <- (plot_b | plot_c) / (plot_d | plot_e) +
  plot_annotation(title = "Diagnostic Plots for Linear Model (2015-2019)")

# Print the combined plot
print(combined_plot)
```
@fig-resid-adj residuals show no clear patterns against attendance, physicians per 1,000, or beds per 1,000, supporting the assumption of linearity. There are no violations against constant variance, unlike in @fig-resid. The fitted values reasonably align with the actual values. However, the limited data points (5 observations) reduce the robustness of these conclusions. 
```{r}
#| echo: false
#| output: false
#| eval: true
#| label: fig-adjmodelresults
#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)."

# Generate individual plots using filtered data (2015-2019)
p1 <- create_projection_plot("attendance", filtered_data, modela)
p2 <- create_projection_plot("physicians_per_1000", filtered_data, modela)
p3 <- create_projection_plot("beds_per_1000", filtered_data, modela)

# Combine the plots into one figure
combined_plot <- (p1 | p2 | p3) +
  plot_annotation(
    title = "Projections of Linear Model (2015-2019) onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
  )

```
Overall, the linear model projections in @fig-adjmodelresults effectively captures general trends, particularly for hospital bed availability. However, the variability around the red lines, particularly for attendance and physicians per 1,000, indicates potential model limitations or missing interactions between predictors. The small sample size (n = 5) further limits the reliability of these insights, necessitating caution when generalizing these findings. Additional data or non-linear modeling may better address these discrepancies and provide a more comprehensive understanding of the relationships between these predictors and wait times.
